Type theories as quotient inductive-recursive types

### 2025-01-25 tex/main.tex
Some progress on the coherence proofs for `π₁,`  in `SC+U+PI+Id.QIIRT`
Fix the coherence proof for `_⇈_`, telescope substitution, and type substitution

\begin{abstract}
  Type theory can be defined in a type theory as a quotient inductive-inductive type, but well-typed terms are littered with explicit coercions, i.e.\ \transp's, along equality constructors and proofs need to take account of coercions painstakingly alongside the interesting part.
  This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with a universe and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
\end{abstract}

@@

\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive types).}
\subsection{Inductive-recursive types}
\cite{Dybjer2003,Dybjer2000,Dybjer1999}

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following equality apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

  To better illustrate the annoyance, note that a complete proof of \eqref{eq:pi2-comp-real-proof} requires us to show each of following equations:
  \begin{alignat*}{3}
         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \\
    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
    = {} & (\sub{\sigma}{\sub{\pi_1\tau}{A}} &&, [\sigma] (\pi_2\,\tau)). \nonumber
  \end{alignat*}
  The first three equations follow from Lemma~2.3.4 in \cite{UFP2013}, while the last equation has nothing to do with substitution calculus but a property of transport (\cite[Lemma~2.3.2]{UFP2013}).
\end{example}

@@

To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor does not compute so that type mismatches occur after substitution.

@@

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching.
In this case, only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
\LT[noinline]{How about \Coq and \Lean?}
Using the first-match semantics, the above definition amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\begin{alignat*}{3}
[ \emptysub ]_{\Ty}\;\UU         & \reduce \UU \\
[ \sigma, t ]_{\Ty}\;\UU         & \reduce \UU \\
[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \reduce \UU,
\end{alignat*}
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.

To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check} which is turned on globally in our formalisation, but we do not trust experimental features blindly.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

To repair the local confluence, we add a simultaneously defined function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
\begin{alignat*}{3}
  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
[ \idS ]_{\Tm}\,t          & \reduce t \\
[ \sigma ; \tau ]_{\Tm}\,t & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
[ \pi_1(\sigma, t) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\,t \\
[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
[ \sigma ]_{\Tm}\,t        & \reduce [ \sigma ]\,t, \quad \text{otherwise}
\end{alignat*}
which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can propagate from type to term substitution.
The remaining cases reduce to the explicit term substitution $[\sigma]\,t$.
Note that $[\blank]_{\Tm}\blank$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.

Then, the equality constructor $[]\El$ and the rule $\mathsf{concat}$ are accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}\\
  \mathsf{concat} : \sigma ; (\tau , t) & =^{\Sub\,\Gamma\,(\Theta, A)} (\sigma ; \tau) , [ \sigma ]_{\Tm} t \nonumber
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \Cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  \LT{clearly terms on the RHS are structurally smaller~\cite{Abel2002}.}
\end{proof}

@@

In this section, we extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, whose the context of its codomain is extended with the domain type.
For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to add the following constructors for the type formation rule, the term introduction/elimination rules:
\begin{alignat*}{3}
  \Pi     &: \implicit{\Gamma}            && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
  \lambda &: \implicit{\Gamma, A, B}      && \Tm\;(\Gamma, A)\;B \to \Tm\,\Gamma\,(\Pi\;A\;B) \\
  \mathsf{app} &: \implicit{\Gamma, A, B} && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
\end{alignat*}
the computation rules
\begin{alignat*}{5}
  \Pi\beta           & : \implicit{\Gamma, A, B, t} && \mathsf{app}\,(\lambda\,t) && =^{\Tm\,\Gamma,A} && t \\
  \Pi\eta            & : \implicit{\Gamma, A, B, t} && t                        && =^{\Tm\,\Gamma,A} && \lambda(\mathsf{app}\,t)
\end{alignat*}
and the structural rules for substitution
\begin{alignat}{5}
  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}    && [ \sigma ]_{\Ty}\,(\Pi\;A\;B) && =^{\Ty\;\Gamma} && \Pi\,(\sub{\sigma}{A})\,([\sigma\uparrow A]\,B) \label{eq:def-type-subst-8} \\
  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t} && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
\end{alignat}
where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \alert{\transfib{\Tm\;\Gamma}{[;]_T}{\color{black}\pi_2\idS}})$.

For our QIIR definition, $\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are definitional.
However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositional but not definitional equal to $\idS$, so the local peak $(\Pi\;A\;B) \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.

>>>>

\begin{abstract}
  Type theory can be defined in a type theory as a quotient inductive-inductive type, but well-typed terms are littered with explicit coercions, i.e.\ \transp's along equality constructors, and proofs need to account for coercions painstakingly alongside the interesting part.
  This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with a universe and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
\end{abstract}

@@

\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive types).}
\subsection{Inductive-recursive types}
\cite{Dybjer2003,Dybjer2000,Dybjer1999}
\LT{Introduce abbreviations here}

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following equality apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ for some $p: [\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\tau]\,A$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

  To better illustrate the annoyance, note that a complete proof of \eqref{eq:pi2-comp-real-proof} requires us to show each of following equations:
  \begin{alignat*}{3}
         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \\
    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
    = {} & (\sub{\sigma}{\sub{\pi_1\tau}{A}} &&, [\sigma] (\pi_2\,\tau)). \nonumber
  \end{alignat*}
  The first three equations corresponding to \eqref{eq:pi2-comp-proof} follow with Lemma~2.3.4 in \cite{UFP2013}, while the last has nothing to do with substitution calculus but the path lifting property (\cite[Lemma~2.3.2]{UFP2013}).
\end{example}

@@

To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor does not compute so that type mismatches occur with substitution.

@@

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching.
In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered.
\LT[noinline]{How about \Coq and \Lean?}
Using the first-match semantics, the above definition amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.

To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
This option is turned on globally in our formal implementation.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

To repair the local confluence, we add a simultaneously defined function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
\begin{alignat*}{3}
  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
[ \idS ]_{\Tm}\,t          & \reduce t \\
[ \sigma ; \tau ]_{\Tm}\,t & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
[ \pi_1(\sigma, t) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\,t \\
[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
[ \sigma ]_{\Tm}\,t        & \reduce [ \sigma ]\,t, \quad \text{otherwise}
\end{alignat*}
which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from type to term substitution.
The remaining cases reduce to the explicit term substitution $[\sigma]\,t$.
Note that $[\blank]_{\Tm}\blank$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.

Then, the equality constructor $[]\El$ is accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \Cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  \LT{clearly terms on the RHS are structurally smaller~\cite{Abel2002}.}
\end{proof}

@@

In this section, we extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, whose the context of its codomain is extended with the domain type.
For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to adding the following constructors for the type formation rule, the term introduction/elimination rules:
\begin{alignat*}{3}
  \Pi     &: \implicit{\Gamma}            && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
  \lambda &: \implicit{\Gamma, A, B}      && \Tm\;(\Gamma, A)\;B \to \Tm\,\Gamma\,(\Pi\;A\;B) \\
  \mathsf{app} &: \implicit{\Gamma, A, B} && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
\end{alignat*}
the computation rules
\begin{alignat*}{5}
  \Pi\beta           & : \implicit{\Gamma, A, B, t} && \mathsf{app}\,(\lambda\,t) && =^{\Tm\,\Gamma,A} && t \\
  \Pi\eta            & : \implicit{\Gamma, A, B, t} && t                        && =^{\Tm\,\Gamma,A} && \lambda(\mathsf{app}\,t)
\end{alignat*}
and the structural rules for substitution
\begin{alignat}{5}
  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}    && [ \sigma ]_{\Ty}\,(\Pi\;A\;B) && =^{\Ty\;\Gamma} && \Pi\,(\sub{\sigma}{A})\,([\sigma\uparrow A]\,B) \label{eq:def-type-subst-8} \\
  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t} && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
\end{alignat}
where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \alert{\transfib{\Tm\;\Gamma}{[;]_T}{\color{black}\pi_2\idS}})$.

For our QIIR definition, $\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are definitional.
However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositional but not definitional equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.

To retain the local confluence, we turn $\sigma \uparrow A$ into a simultaneously defined function:
\begin{alignat}{3}
  \blank\uparrow \blank & : (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \label{eq:type-of-lifting} \\
\idS                \uparrow A & \reduce \idS \nonumber \\
(\sigma ; \tau)     \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \nonumber \\
\pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \nonumber \\
\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow \sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A)\nonumber \\
\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwise} \nonumber
\end{alignat}
where $\sigma^+ \defeq (\pi_1\idS ; \sigma, \pi_2\idS)$.
Then, we add another clause for type substitution and an equality constructor for term substitution without transport:
\begin{alignat}{5}
                     & [ \sigma ]_{\Ty}\,(\Pi\,A\,B) && \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \label{eq:def-type-subst-9} \\
  \mathsf{[]\lambda} & : [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]_{\Tm}\,t) \nonumber
\end{alignat}
